\demes: a pain-free specification for complex demography

### 2021-04-30 paper.tex

\title{\demes: a Python package for pain-free specification of complex demography}
\author[1,*]{Graham Gower}
\author[2,*]{Jerome Kelleher}
\author[3,*]{Aaron P. Ragsdale}
\author[4,*]{Kevin Thornton}
\affil[1]{}
\affil[2]{}
\affil[3]{}
\affil[4]{}
\affil[*]{Authors listed alphabetically}
\maketitle

\abstract{
come back to title and abstract later
}

@@

Here, we present \demes, a Python package for specifying complex demographic
models in a way that is both human readable and can be directly passed to a
growing number of simulation engines. \Demes\ supports model specification in a
high-level \YAML\ format \citep{ben2009yaml} that is designed to be as easy as
possible to implement, read, and share. Models are then represented in an
unambiguous low-level format within Python that checks model validity and can
then be used as the input for simulation software.  The initial release of
\demes\ (version 0.1, described here) supports \emph{static} demographic
descriptions, that is, models with fixed parameters. Future releases will
support [drawing parameters from distributions, for example for ABC-based
inference, or describing ``dynamic`` models where parameters may be fit to
data, e.g. \moments\ or \dadi].

\section*{Defining demographic models with \demes}

@@

\begin{itemize}
\item Short description of YAML language and why it's appropriate here
\item General overview (1 paragraph) of defining demography in YAML
\item A generic, commonly used, simple demographic model for illustration (e.g.
  an IM in Model~\ref{code:im_model})
\end{itemize}

@@

\subsection*{Python representation of demography as input for simulation software}

\begin{itemize}
\item The Python API, reading in a YAML, and unambiguous and manipulable representation
\end{itemize}

\subsubsection*{Software support for \demes}
\aprcomment{at time of publication. TBD...}
\begin{itemize}
\item \msprime
\item \stdpopsim
\item \fwdpy
\item \moments
\item \dadi? should try to get Ryan on board
\item \slim? I guess that comes with \stdpopsim
\end{itemize}

\subsection*{Specifying complex demography}

\aprcomment{A few examples of complex demography - e.g. 1) the very common out-of-Africa
model, 2) Denisovan admixture with a ton of populations and demographic events, 3) a
non-human model, perhaps with selfing rates and rate changes}

\aprcomment{Perhaps examples should be in the appendix, to keep the paper concise}

\section*{Prospects and future directions}

@@

>>>>

\title{\demes: a pain-free specification for complex demography}
\author[*]{Graham Gower}
\author[*]{Aaron P. Ragsdale}
\author[ ]{...Others here...}
\author[**]{Kevin Thornton}
\author[**]{Jerome Kelleher}
\affil[*,**]{Denote equal contribution}
\maketitle

\abstract{
}

@@

Here, we present \demes, a specification of complex demographic models that is
both human readable and can be directly passed to a growing number of
simulation software. \Demes\ supports model specification in a high-level
\YAML\ format \citep{ben2009yaml} that is designed to be as easy as possible to
implement, read, and share. Models are then represented in an unambiguous
low-level format within Python that checks model validity. The initial Python
release of \demes\ (version 0.1, described here) supports \emph{static}
demographic descriptions, that is, models with fixed parameters. Future
releases will support [[drawing parameters from distributions, for example for
ABC-based inference, or describing ``dynamic`` models where parameters may be
fit to data, e.g. \moments\ or \dadi]].

\section*{The \demes\ specification}

\begin{itemize}
    \item Demographic models are comprised of one or more interacting
        populations, along with description, doi, time units and generation
        times.
    \item Populations (or demes) are treated as discrete collection of
        exchangeable individuals following a well-defined set of rules
    \item These rules include population sizes, times of existence, and other
        information about the mating system (e.g. selfing or cloning rates)
    \item Relationships between demes are defined by ancestors/descendents
        and possible continuous or instantaneous migration between coexisting
        demes
\end{itemize}

@@

\begin{itemize}
    \item YAML allows us to write this information in a human-readable and
        structured manner (lists, keys/items, etc)
    \item Implicit values, so that models can be written compactly
    \item Example of an IM Model~\ref{code:im_model}
\end{itemize}

@@

\subsection*{Unambiguous low-level representation}

\begin{itemize}
    \item The Python API, which can read in a YAML file or be built using the
        Builder
    \item Demographic model validitation
    \item Unambiguous, redundant information
    \item Designed to be programatically read by simulation software backends
\end{itemize}

\section*{Extensions and future directions}

\subsection*{Software support for \demes}
\aprcomment{at time of publication. TBD...}
\begin{itemize}
    \item \msprime: coalescent simulation \citep{kelleher2016efficient}
    \item \stdpopsim: a collection of maintained species and demographic models,
        which includes a large number of models specified using \demes
        \citep{adrion2020community}
    \item \fwdpy: forward-time Wright-Fisher simulation \citep{thornton2014c++}
    \item \moments: compute expected diversity statistics (SFS, LD) and perform
        demographic inference using summary statistics
        \citep{jouganous2017inferring,ragsdale2019models}
    \item \dadi: compute the expected SFS and perform demographic inference using
        summary statistics \citep{gutenkunst2009inferring}
\end{itemize}

\subsection*{Other stuff...}

\begin{itemize}
    \item \texttt{demesdraw} (\url{https://github.com/grahamgower/demesdraw.git})
    \item General framework for inference (have been working on that over in \moments,
        and perhaps it could be generalized to work with any simulation engine
    \item Defining distributions of parameter values, perhaps?
\end{itemize}

@@

\aprcomment{Perhaps and example of a many-population model, and API to simulate using
\msprime, \fwdpy, and \moments?}
